Exploring probability distributions for bivariate temporal granularities

### 2019-09-15 main.Rmd
section on single and multiple order granularity  updated example of  mayan included

<!-- What is multiple and single order up granularities -->
The hierarchical structure of time creates a natural nested ordering, where hours are nested within days, days within weeks, weeks within months, and so on. We refer to granularities which are nested within multiple levels as **multiple-order-up** granularities. For example, hour of the week and second of the hour are both multiple-order-up, while hour of the day and second of the minute are **single-order-up**.

@@

| Temporal units	| Conversion factor 	| conversion in days 	| single-order-up accessors 	|
|------------	|-------------------	|--------------------	|---------------------------	|
| 1 kin      	| 1 day             	|                    	| kin_uinal                 	|
| 1 uinal    	| 20 kin            	| 20 days            	| uinal_tun                 	|
| 1 tun      	| 18 uinal          	| 360 days           	| tun_katun                 	|
| 1 katun    	| 20 tun            	| 7200 days          	| katun_baktun              	|
| 1 baktun   	| 20 katun          	| 144,000 days       	| 1                         	|

@@

@@

**Definition**: A **hierarchy table** has three columns:  
 - The first column represents the linear granularities in ascending order of temporal hierarchy.  
 - The second column represents the constant which relates subsequent linear granularities.  
 - The third column provides the accessor function for single-order up circular granularities.

@@

sm_cust50 %>%
  create_gran("day_week") %>%
  filter(day_week %in% c("Sat", "Sun")) %>%
  granplot("day_week", "week_month",
           response = "general_supply_kwh",
           plot_type = "lv") + 
  ggtitle("") +
  ggtitle("Letter value plot across day-of-week by week-of-month") +
  scale_y_continuous(breaks = seq(0, 1.5, 0.25))
```

@@

cricket_tsibble %>% granplot("ball_over", "over_inning",
  hierarchy_model,
  response = "total_runs",
  plot_type = "quantile",
  quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))
)

@@

cricket_tsibble %>% granplot("ball_over", "over_match",
  hierarchy_model,
  response = "total_runs"
) + scale_color_brewer(type = "div", palette = "Dark2")

@@

The scores/ball increases from 1st to 120th ball in the 90th percentile. In the last 2 overs, players are more vulnerable to get more scores, which is evident from the fact that in 90 percent of cases, their scores vary between 4 and 6. Till around 50th ball, players play safe and may get no runs per ball in 25% of the times, after which they more likely to get one score per ball.

<!-- # ```{r over, echo=TRUE} -->
<!-- # data_MI %>% ggplot(aes(x = as.factor(over), y= total_runs)) + ggtitle("Letter value plot of runs per ball across each over of a match") + geom_lv(aes(fill = ..LV..), outlier.colour = "red", outlier.shape = 1, k = 5) + xlab("overs per match") -->

<!-- # data_MI %>% ggplot(aes(y =as.factor(over), x= total_runs)) + geom_density_ridges2() + ylab("Balls per  over") -->
<!-- ``` -->

<!-- In the first two overs, players play really safe and make no runs 50% of the times. The median is constant at 1 for all overs after the 2nd over. It is interesting to see the distribution of tails in this plot. M, F, E, D and C represents 50%, 25%, 12.5%, 6.25% and 3.13% of the tail area respectively. Till the first four overs, scoring a 4 or a 6 is extreme. The region which covers the tail 3.13% falls between 4 and 6 till up to 13th over. It is interesting to see how the letter values of scores (D, E and F) moves up, as we move towards the end of the game.  -->

<!-- ```{r quarter, echo=TRUE} -->
<!-- data_MI %>% ggplot(aes(x = as.factor(quarter_match), y= total_runs)) +  geom_violin()+ xlab("quarters per match")+ ylab(" Total runs per ball") + ggtitle("Violin plot of runs per ball across each quarter of a match") -->

<!-- # data_MI %>% ggplot(aes(y =as.factor(quarter), x= total_runs)) + geom_density_ridges2() + ylab("quarters per match") + xlab(" Total runs per ball") -->
<!-- ``` -->

<!-- 4th quarter of the match witnesses low density(mass) at scores 0, 1 and more at 4/6. The density at score 6 increases with the quarters of the match. -->

<!-- ```{r semester, echo=TRUE} -->
<!-- # data_MI %>% ggplot(aes(x = as.factor(semester_match), y= total_runs)) +  geom_violin()+ xlab("semester per match")+ ylab(" Total runs per ball") +ggtitle("Violin plot of runs per ball across each semesters of a match") -->

<!-- data_MI %>% ggplot(aes(x = total_runs, y= as.factor(semester_match))) +  geom_density_ridges2()+ ylab("semesters per match")+ xlab(" Total runs per ball")+ ggtitle("Ridge plot of runs per ball across each semesters of a match") -->

<!-- # data_MI %>% ggplot(aes(y =as.factor(semester), x= total_runs, fill=factor(..quantile..))) +   stat_density_ridges( -->
<!-- #     geom = "density_ridges_gradient", calc_ecdf = TRUE, -->
<!-- #     quantiles = 4, quantile_lines = TRUE -->
<!-- #   )  + ylab("semester") + xlab(" Total runs per ball") + -->
<!-- #   scale_fill_viridis(discrete = TRUE, name = "Quartiles") -->
<!-- ``` -->

<!-- In the second semester of the game, there is more probability to score more than zero as can be seen from the drop in peaks of the distribution at 0 and increased peak at 1, 2, 4 and 6. -->

<!-- Now, it will be interesting to see the distribution of runs per ball across two categorisation together: -->

<!-- ```{r bivariate1, echo=TRUE} -->
<!-- # data_MI %>% ggplot(aes(x = as.factor(semester_match), y= total_runs)) +  geom_violin()+ xlab("semester per match")+ ylab(" Total runs per ball") +ggtitle("Violin plot of runs per ball across each semesters of a match") -->

<!-- data_MI %>% ggplot(aes(x = as.factor(over_quarter), y = total_runs))+  geom_violin()+ -->
<!--                      facet_wrap(~quarter_semester) + xlab("overs per quarter")+ ylab(" Total runs per ball") + ggtitle("Boxplot of runs per ball across each over_quarter and quarter_semester")  -->
<!-- ``` -->

<!-- # runs per over - as variation would be more -->

<!-- ```{r checking} -->
<!-- data_MI_over <- data_MI %>% group_by(match_id, over) %>% summarise(runs_over = sum(total_runs)) %>%  -->
<!--     mutate(quarter_match = case_when( -->

<!--     between(over,1,5) ~ 1, -->
<!--     between(over,6,10) ~ 2, -->
<!--     between(over,11,15) ~ 3, -->
<!--     between(over,16,20) ~ 4 -->
<!--   )) %>%  -->
<!--    mutate(semester_match = case_when( -->

<!--     between(quarter_match,1,2) ~ 1, -->
<!--     between(quarter_match,3,4) ~ 2 -->
<!--   ))  %>%  -->
<!--    mutate(over_quarter =  -->
<!--   if_else(over%%5==0,5, over%%5) -->
<!--    ) %>%  mutate(quarter_semester =  -->
<!--   if_else(quarter_match%%2==0,2, quarter_match%%2) -->
<!--    ) %>%  -->
<!--   rename(over_match = over) -->

<!-- ``` -->

<!-- ```{r over_match, echo=TRUE} -->
<!-- data_MI_over %>% ggplot(aes(x = as.factor(over_match), y= runs_over)) + ggtitle("Letter value plot of runs per over across each over of a match") + geom_lv(aes(fill = ..LV..), outlier.colour = "red", outlier.shape = 1, k = 5) + xlab("overs per match") -->

<!-- # data_MI %>% ggplot(aes(y =as.factor(over), x= total_runs)) + geom_density_ridges2() + ylab("Balls per  over") -->
<!-- ``` -->

<!-- ```{r over_quarter, echo=TRUE} -->
<!-- data_MI_over %>% ggplot(aes(x = as.factor(over_quarter), y= runs_over)) +  geom_violin()+ xlab("quarters per match")+ ylab(" Total runs per ball") + ggtitle("Violin plot of runs per over across each quarter of a match") -->

<!-- data_MI_over %>% ggplot(aes(x = as.factor(over_quarter), y= runs_over)) +  geom_violin()+ xlab("quarters per match")+ ylab(" Total runs per ball") + ggtitle("Violin plot of runs per over across each quarter of a match") + facet_grid(~quarter_match) -->

<!-- # data_MI %>% ggplot(aes(y =as.factor(quarter), x= total_runs)) + geom_density_ridges2() + ylab("quarters per match") + xlab(" Total runs per ball") -->
<!-- ``` -->

<!-- ```{r quarter_match, echo=TRUE} -->
<!-- # data_MI %>% ggplot(aes(x = as.factor(semester_match), y= total_runs)) +  geom_violin()+ xlab("semester per match")+ ylab(" Total runs per ball") +ggtitle("Violin plot of runs per ball across each semesters of a match") -->

<!-- data_MI_over %>% ggplot(aes(x = runs_over, y= as.factor(quarter_match))) +  geom_density_ridges2()+ ylab("semesters per match")+ xlab(" Total runs per ball")+ ggtitle("Ridge plot of runs per overs across each semesters of a match") -->

<!-- data_MI_over %>% ggplot(aes(x = runs_over, y= as.factor(over_quarter))) +  geom_density_ridges2()+ ylab("semesters per match")+ xlab(" Total runs per ball")+ ggtitle("Ridge plot of runs per overs across each semesters of a match") + facet_wrap(~quarter_match) -->

<!-- # data_MI %>% ggplot(aes(y =as.factor(semester), x= total_runs, fill=factor(..quantile..))) +   stat_density_ridges( -->
<!-- #     geom = "density_ridges_gradient", calc_ecdf = TRUE, -->
<!-- #     quantiles = 4, quantile_lines = TRUE -->
<!-- #   )  + ylab("semester") + xlab(" Total runs per ball") + -->
<!-- #   scale_fill_viridis(discrete = TRUE, name = "Quartiles") -->
<!-- ``` -->

>>>>

<!-- What is multiple and single order up granularities -->
The hierarchical structure of time creates a natural nested ordering. Hence, given a hierarchy we can define **single-order-up** and **multiple-order-up** granularities. We refer to granularities which are nested within multiple levels as **multiple-order-up** granularities. We take the help of Mayan calendar to illustrate this. In "Mayan" calendar, one day was referred to as 1 kin and the calendar was structured as follows:

| Units	      | Conversion factor
|------------	|-------------------	|
|  kin      	| 20             	    |
|  uinal    	| 18            	    |
|  tun      	| 20           	      |
|  katun    	| 20            	    |
|  baktun   	| 1          	        |

Examples of **multiple-order-up**  granularities can be kin of the tun or kin of the baktun whereas examples of single-order-up granularities, which are nested within single levels may include kin of the uinal, uinal of the tun etc.

In the next section, we discuss the computation of any single-order-up and multiple-order-up granularities in a periodic and aperiodic set up.

@@

| Units	      | Conversion factor
|------------	|-------------------	|
|  index      | 1                   |
|  kin      	| 20             	    |
|  uinal    	| 18            	    |
|  tun      	| 20           	      |
|  katun    	| 20            	    |
|  baktun   	| 1          	        |

@@

Suppose, the index of the tsibble is

@@

**Definition**: A **hierarchy table** has three columns:  
 - The first column represents the linear granularities in ascending order of temporal hierarchy.  
 - The second column represents the constant which relates subsequent linear granularities.

@@

sm_cust50 %>%
  dynamic_create_gran("day_week") %>%
  filter(day_week %in% c("Sat", "Sun")) %>%
  granplot("day_week", "week_month",
           response = "general_supply_kwh",
           plot_type = "lv") + 
  ggtitle("") +
  ggtitle("Letter value plot across day-of-week by week-of-month") +
  scale_y_continuous(breaks = seq(0, 1.5, 0.25))

sm_cust50 %>%
  granplot("fortnight_month", "hour_day",
           response = "general_supply_kwh",
           plot_type = "lv") + 
  ggtitle("")
```

@@

cricket_tsibble %>% granplot("ball_over", "over_inning",
  hierarchy_model,
  response = "total_runs",
  plot_type = "quantile",
  quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9))

@@

cricket_tsibble %>% granplot("ball_over", "inning_match",
  hierarchy_model,
  response = "total_runs"
) + scale_color_brewer(type = "div", palette = "Dark2")

@@

To see more variation in the response variable, we can also see the distribution of runs per over.

```{r runs_per-over}
cricket_per_over <- cricketdata %>% 
  group_by(match_id, batting_team, bowling_team,  inning, over) %>%
  summarise(runs_per_over = sum(total_runs))

cricket_tsibble <- cricket_per_over %>% 
  ungroup() %>% 
  mutate(data_index = row_number()) %>% 
  as_tsibble(index = data_index)

  cricket_tsibble %>% granplot("inning_match", "over_inning",
                               hierarchy_model,
                               response = "runs_per_over",
                               plot_type = "quantile",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)

  cricket_tsibble %>%
    filter(batting_team == "Chennai Super Kings") %>% 
    granplot("inning_match", "over_inning",
                               hierarchy_model,
                               response = "runs_per_over",
                               plot_type = "quantile",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)
  )

    cricket_tsibble %>%
    filter(batting_team == "Kolkata Knight Riders") %>% 
    granplot("inning_match", "over_inning",
                               hierarchy_model,
                               response = "runs_per_over",
                               plot_type = "quantile",
                               quantile_prob = c(0.1, 0.25, 0.5, 0.75, 0.9)
  )
